---
title: "MS brain age UCL MAGNIMS analysis"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
---
James Cole - October 2019. Built with R version `r getRversion()`  
This is Notebook contains the final brain age analysis of MS patient data and controls from the UCL cohort, the MAGNIMS consortium and the Imperial College London PET study (n=25). The analysis uses brain-predicted age difference (brain-PAD) to look at brain ageing in the context of MS. The brain-PAD values were generated in PRONTO, using an independent healthy (n=2001) training dataset, and the values were corrected for proportional bias using the intercept and slope of the age by brain-predicted age regression in the training dataset.
  
### Initial set up of analysis
Clear workspace, load libraries, set colour palette
```{r message=FALSE, warning=FALSE}
rm(list = ls()) ## clear workspace
library(car)
library(corrplot)
library(cowplot)
library(data.table)
library(dplyr)
library(emmeans)
library(ggstance)
library(ggplot2)
library(grDevices)
library(gridExtra)
library(hier.part)
library(jtools)
library(knitr)
library(lmerTest)
library(MASS)
library(metafor)
library(plotrix)
library(plyr)
library(pryr)
library(psych)
library(qwraps2)
library(reshape)
library(reshape2)
library(RColorBrewer)
library(scales)
library(stringr)
library(survival)
library(survminer)
library(tidyr)
ms.palette <- c("darkgreen", "darkorange", "red", "blue", "purple") # define MS colour scheme for groups
sessionInfo()
```
#### Get data from CSV and define longitudinal data.frames
```{r}
setwd('/Users/jcole/Work/Brain ageing/Collaborations/MS')
df <- read.csv("MS_brain_age_final_long.csv")
df$subtype <- factor(df$subtype, levels = c("control", "CIS", "RRMS", "SPMS", "PPMS")) # reorder subtype factor to put controls first
df$Cohort <- recode(df$Cohort, JR1 = "Imperial", C0 = "UCL0", C1 = "UCL1", C2 = "UCL2", C3 = "UCL3", C4 = "UCL4" ,C5 = "UCL5", C6 = "UCL6", C7 = "UCL7", A = "Amsterdam", B = "Barcelona", G = "Graz", M = "Milan", R = "Rome", S = "Siena") 
```
#### Exclude participants with errors in the database & correct time since diagnosis errors
```{r message=FALSE, warning=FALSE}
tmp <- df[df$Cohort == 'Amsterdam',] %>% group_by(PatientID) %>% dplyr::summarize(sd = sd(age_at_scan)) %>% arrange(desc(sd)) %>% filter(sd > 2)
excluded_IDs <- sort(tmp$PatientID)
# excluded_IDs <- unique(df[which(df$age_at_scan < df$age_at_baseline_scan1),1])
excluded_scans <- (df %>% filter(str_detect(PatientID, paste(excluded_IDs, collapse = "|"))))["ScanName"]
df <- df %>%
  filter(!str_detect(PatientID, paste(excluded_IDs, collapse = "|")))
rm(tmp)
```
There were `r length(excluded_IDs)` subjects with `r dim(excluded_scans)[1]` scans excluded in total. Data entry errors in original spreadsheet; age at baseline not consistent within subject.

Load data for lesion filling analysis
```{r}
## read in data
lesion_df <- read.csv(file = '~/Work/Brain ageing/Collaborations/MS/MAGNIMS/MAGNIMS20171224_final.csv')
lesion_df$subtype <- factor(lesion_df$subtype, levels = c("control", "CIS", "RRMS", "SPMS", "PPMS"))
## create ICV ratio variables
lesion_df <- lesion_df %>%
  mutate(ICV = GMV + WMV + CSFV) %>%
  mutate(gm_icv_ratio = GMV/ICV) %>%
  mutate(wm_icv_ratio = WMV/ICV)
```

#### Generate baseline only data.frame and show data frame structure
```{r}
df.bl <- df[(df$age_at_baseline_scan1 == df$age_at_scan),] # define baseline data.frame
str(df)
```
### Basic stats
#### Total number of subjects, and by group
The total number of subjects included was n = `r dim(df.bl)[1]`

The total number of MS patients (including CIS) was  n = `r table(df.bl$control0rest1)[2]` and healthy controls was n = `r table(df.bl$control0rest1)[1]`

#### Number of scans in total and per group
Total number of scans = `r sum(df.bl$NoScans)`


#### Number of people with 2 or more scans
```{r}
df.bl %>%
  filter(NoScans >= 2) %>%
  group_by(control0rest1) %>%
  tally()
```
#### Number of people with 3 or more scans
```{r}
df.bl %>%
  filter(NoScans >= 3) %>%
  group_by(control0rest1) %>%
  tally()
```
### Generate demographics table using qwarps2
```{r message=FALSE, warning=FALSE}
options(qwraps2_markup = "markdown", digits = 2)

table1 <-
  list("N" =
         list("Control" = ~ qwraps2::n_perc0(control0rest1 == "control", na_rm = T),
              "MS"  = ~ qwraps2::n_perc0(control0rest1 == "MS", na_rm = T)),
     "Gender" =
       list("Female" = ~ qwraps2::n_perc0(gender == "female", show_symbol = T),
            "Male"  = ~ qwraps2::n_perc0(gender == "male", show_symbol = T)),
        "Number of scans" =
       list("min" = ~ min(NoScans),
            "max" = ~ max(NoScans),
            "mean (sd)" = ~ qwraps2::mean_sd(NoScans)),
       "Age at baseline scan (years)" =
       list("min" = ~ min(age_at_baseline_scan1),
            "max" = ~ max(age_at_baseline_scan1),
            "mean (sd)" = ~ qwraps2::mean_sd(age_at_baseline_scan1)),
       "Brain-predicted age at baseline scan (years)" =
       list("min" = ~ min(BrainAge),
            "max" = ~ max(BrainAge),
            "mean (sd)" = ~ qwraps2::mean_sd(BrainAge)),
       "Disease duration at baseline (years)" =
       list("min" = ~ min(disease_duration, na.rm = T),
            "max" = ~ max(disease_duration, na.rm = T),
            "mean (sd)" = ~ qwraps2::mean_sd(disease_duration, na_rm = T, show_n = "never")),
        "EDSS at baseline " =
       list("min" = ~ min(EDSSbaseline, na.rm = T),
            "max" = ~ max(EDSSbaseline, na.rm = T),
            "mean (sd)" = ~ qwraps2::mean_sd(EDSSbaseline, na_rm = T, show_n = "never")),
       "MS subtype" =
       list("CIS" = ~ qwraps2::n_perc0(subtype == "CIS", show_symbol = T),
            "RRMS"  = ~ qwraps2::n_perc0(subtype == "RRMS", show_symbol = T),
            "SPMS"  = ~ qwraps2::n_perc0(subtype == "SPMS", show_symbol = T),
            "PPMS"  = ~ qwraps2::n_perc0(subtype == "PPMS", show_symbol = T)),
       "Treatment" =
         list("Yes" = ~ qwraps2::n_perc0(DMT_YesNoNA == "YES", na_rm = T, show_symbol = T),
              "No"  = ~ qwraps2::n_perc0(DMT_YesNoNA == "NO", na_rm = T, show_symbol = T))
       )

print(summary_table(dplyr::group_by(df.bl, control0rest1), table1),
      rtitle = "Sample Chararcteristics",
      cnames = c("Controls", "MS patients"))
```
#### Need to get treatment NAs using table()
```{r}
table(df.bl$DMT_YesNoNA, df.bl$control0rest1, useNA = "ifany")[,"MS"]
```

#### Need to get length of follow-up from longitudinal database
##### Length of follow-up
```{r}
options(digits = 3) ## return digits option to default
df %>%
  filter(NoScans >= 2) %>%
  group_by(PatientID) %>%
  slice(which.max(interval)) %>%
  # top_n(n = 1, wt = interval) %>%
  group_by(control0rest1) %>%
  dplyr::summarise(mean(interval), sd(interval), min(interval), max(interval))
options(digits = 7) ## return digits option to default
```

Correlations between demographics and clinical variables
```{r}
cor.mat <- cbind(df.bl$age_at_scan, df.bl$disease_onset_age, df.bl$disease_duration, df.bl$EDSSatScan)
colnames(cor.mat) <- c("Age at scan", "Age at diagnosis", "Time since diagnosis", "EDSS at scan")
corrplot(cor(cor.mat, use = "pairwise"), type = "upper", method = "color", addCoef.col = T, tl.col = "black", diag = F)
```

### Baseline brain-age analysis
```{r paged.print=FALSE}
describeBy(df.bl$BrainPAD, df.bl$control0rest1, mat = T, digits = 3) # brain-PAD by MS patient vs. controls
```
#### Estimated marginal means
Generate EMMs for all MS/CIS and healthy controls. LME adjusting for age, gender, ICV, cohort and scanner status.
```{r}
fit <- (lmer(BrainPAD ~ control0rest1 + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = df.bl))
emmeans(object = fit, specs = c("control0rest1", "age_at_baseline_scan1", "ICV"))
```

#### Effects of cohort and scanner status on brain-PAD
The significant interaction motivates nesting in the LME.
```{r}
with(df.bl, anova(lm(BrainPAD ~ Cohort * scanner_status)))
```
#### Meta-analysis looking at all the separate cohorts with MS/CIS patients and controls
Check which cohorts contain healthy controls and patients.
```{r}
table(df.bl$subtype, df.bl$Cohort)
```

Create data.frame with summary data appropriate for meta-analysis.
```{r}
tmp0 <- df.bl %>% group_by(Cohort, control0rest1) %>% dplyr::summarise(length(control0rest1))
meta.cohorts <- as.list(tmp0[tmp0$control0rest1 == "control",1])$Cohort
tmp <- df.bl %>%
  filter(str_detect(Cohort, paste(meta.cohorts, collapse = "|")))
tmp1 <- tmp %>% group_by(Cohort, control0rest1) %>% dplyr::summarise(n = n(), Mean = mean(BrainPAD), SD  =  sd(BrainPAD))
tmp2 <- dcast(tmp1, Cohort ~ control0rest1, value.var = "n")
tmp3 <- dcast(tmp1, Cohort ~ control0rest1, value.var = "Mean")
tmp4 <- dcast(tmp1, Cohort ~ control0rest1, value.var = "SD")
names(tmp2) <- c("Cohort", "control_n", "MS_n")
names(tmp3) <- c("Cohort", "control_mean", "MS_mean")
names(tmp4) <- c("Cohort", "control_sd", "MS_sd")
tmp1x <- tmp %>% group_by(Cohort) %>% dplyr::summarise(pooled.SD  =  sd(BrainPAD))
meta.df <- join_all(list(tmp2,tmp3,tmp4, tmp1x), by = 'Cohort', type = 'left')
rm(list = ls(pattern = 'tmp*')) # remove temporary data frames
```
Run meta-analysis using the metafor package, to fit a random-effects meta-analysis using REML.
```{r}
meta.df <- escalc(m1i = control_mean, sd1i = pooled.SD, n1i = control_n, m2i = MS_mean, sd2i = pooled.SD, n2i = MS_n, measure = "MD", data = meta.df, digits = 2)
meta.results <- rma(yi, vi, data = meta.df, method = "REML")
print(meta.results)
confint(meta.results)
```
Forest plot of results
```{r message=FALSE, warning=FALSE}
plot.forest %<a-% {
forest(meta.results, ilab = cbind(meta.df$MS_n, meta.df$control_n), ilab.xpos = c(-30,-23), slab = meta.df$Cohort, digits = 1, xlab = "MS vs. Healthy control group mean difference", steps = 6, col = "red", cex = 1.25, pch = 22, bg = "blue"); text(c(-40, -30, -23), 7.6, c("Cohort", "MS n", "HC n"), font = 2, cex = 1.25)
}
plot.forest
cairo_pdf("plots/forest_plot.pdf", 6,5)
plot.forest
dev.off()
```

Linear regression analysis restricted to cohort UCL3, includes covariates: age, gender, intracranial volume.
```{r}
summary(lm(BrainPAD ~ control0rest1 + age_at_baseline_scan1 + gender + ICV, data = subset(df.bl, df.bl$Cohort == "UCL3")))
```
### Main result:
#### LME model to predict brain-PAD based on group
```{r}
# LME model of control vs. MS, including Cohort and scanner status as random effects
m1 <- lmer(BrainPAD ~ control0rest1 + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = df.bl)
summary(m1)
round(confint(m1),2)
```

#### Lesion filling
To establish whether using the FSL lesion filling software influences brain-predicted age values.
This analysis was conducted only in UCL patients. 
```{r paged.print=FALSE}
with(subset(lesion_df, lesion_df$subtype != "control"), describeBy(filled_brain_age, subtype))
```
Correlation between brain-predicted age from filled and unfilled images: Pearson's r = `r round(with(subset(lesion_df, lesion_df$subtype != "control"), cor(filled_brain_age, brain_age, use = "pairwise")),3)`.
Median absolute error (MAE) between brain-predicted age from filled and unfilled images = `r with(subset(lesion_df, lesion_df$subtype != "control"), median(abs(filled_brain_age - brain_age), na.rm = T))` years.
Mean difference between brain-predicted age from filled and unfilled images = `r round(with(subset(lesion_df, lesion_df$subtype != "control"), mean(filled_brain_age - brain_age, na.rm = T)),3)` years.

```{r}
ggplot(data = subset(lesion_df, lesion_df$subtype != "control"), aes(x = brain_age, y = filled_brain_age)) +
  geom_abline(slope = 1) +
  geom_point(pch = 21, aes(fill = subtype), size = 2) +
  labs(x = "Unfilled brain-predicted age (years)", y = "Lesion-filled brain-predicted age (years)") +
  xlim(c(20,90)) +
  scale_fill_manual(values = ms.palette[-1]) +
  theme_bw() + theme(legend.position = c(0.9, 0.2))
ggsave("~/Work/Brain ageing/Collaborations/MS/plots/lesion_filling_brain_age_plot.pdf", width = 5, height = 5, useDingbats = FALSE)
```
#### Bland-Altman plot
```{r}
mean.diff <- mean(lesion_df$brain_age - lesion_df$filled_brain_age, na.rm = T)
sd.diff <- sd(lesion_df$brain_age - lesion_df$filled_brain_age, na.rm = T)
ggplot(data = subset(lesion_df, lesion_df$subtype != "control"), aes(x = ((brain_age + filled_brain_age)/2), y = brain_age - filled_brain_age)) +
  geom_abline(slope = 0, lty = 2) +
  geom_point(pch = 21, aes(fill = subtype), size = 2) +
  geom_hline(yintercept = mean.diff, color = "darkgoldenrod1", lwd = 1) + # mean difference line
  geom_hline(yintercept = mean.diff + 1.96*sd.diff, color = "darkgoldenrod2", lty = 2) + # upper 95% line
  geom_hline(yintercept = mean.diff - 1.96*sd.diff, color = "darkgoldenrod2", lty = 2) + # lower 95% line
  # geom_smooth(method = "lm", level = 0.95, color = "black", lwd = 0.3) +
  labs(x = "Mean of filled/unfilled brain-predicted age (years)", y = "Unfilled - Lesion-filled brain-predicted age (years)") +
  # ylim(c(-20,20)) +
  scale_fill_manual(values = ms.palette[-1]) +
  theme_bw() + theme(legend.position = c(0.1, 0.8)) +
  annotate("text", x = 25, y = mean.diff + 1.96*sd.diff + 0.5, label = "+1.96*SD", color = "darkgoldenrod2") +
  annotate("text", x = 25, y = mean.diff - 1.96*sd.diff - 0.5, label = "-1.96*SD", color = "darkgoldenrod2") +
  annotate("text", x = 27.5, y = mean.diff + 0.8, label = "Mean difference", color = "darkgoldenrod2")
ggsave("~/Work/Brain ageing/Collaborations/MS/plots/lesion_filling_brain_age_BA_plot.pdf", width = 8, height = 5, useDingbats = FALSE)
```

#### LME model predicted brain-PAD based on subtype
This analysis excluded controls.
```{r}
# LME model of subtype, including Cohort and scanner status as random effects
m2 <- lmer(BrainPAD ~ subtype + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "control"))
summary(m2)
# round(confint(m2),2)
```
#### Brain-PAD estimated marginal means for subtypes
Generate EMMs for all MS subtypes. LME adjusting for age, gender, ICV, cohort and scanner status.
```{r}
fit <- (lmer(BrainPAD ~ subtype + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "control")))
emmeans(object = fit, specs = c("subtype", "age_at_baseline_scan1", "ICV"))
```

### Brain-PAD boxplot by MS subtype
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# calculate Ns
control_n <- with(df.bl, table(subtype))["control"]
cis_n <- with(df.bl, table(subtype))["CIS"]
rrms_n <- with(df.bl, table(subtype))["RRMS"]
spms_n <- with(df.bl, table(subtype))["SPMS"]
ppms_n <- with(df.bl, table(subtype))["PPMS"]

plot.pryr %<a-% {
with(df.bl, ehplot(BrainPAD, groups = subtype, ylim = c(-20,52), pch = 21, bg = ms.palette[as.numeric(subtype)], box = T, offset = 0.1, intervals = 50, xlab = "Groups", ylab = "Brain-PAD (years)", main = "Baseline Brain-PAD by study group")) + 
  abline(0,0, lty = 2) +
  text(1,51, paste("N = ", control_n, sep = ""), cex = 0.8) +
  text(2,51, paste("N = ", cis_n, sep = ""), cex = 0.8) +
  text(3,51, paste("N = ", rrms_n, sep = ""), cex = 0.8) +
  text(4,51, paste("N = ", spms_n, sep = ""), cex = 0.8) +
  text(5,51, paste("N = ", ppms_n, sep = ""), cex = 0.8)
}
plot.pryr
cairo_pdf("plots/MS_UCL_MAGNIMS_combined_group_Brain-PAD.pdf", 8,8)
plot.pryr
dev.off()
```
### Brain-PAD boxplot by MS subtype in cohort UCL3 only
```{r message=FALSE, warning=FALSE}
# calculate Ns
C3.bl <- df.bl %>% filter(Cohort == "UCL3") 
C3.bl$subtype <- factor(C3.bl$subtype)
control_n <- with(C3.bl, table(subtype))["control"]
rrms_n <- with(C3.bl, table(subtype))["RRMS"]
spms_n <- with(C3.bl, table(subtype))["SPMS"]
ppms_n <- with(C3.bl, table(subtype))["PPMS"]

plot.pryr %<a-% {
  with(C3.bl, ehplot(BrainPAD, groups = subtype, ylim = c(-20,52), pch = 21, bg = ms.palette[-2][as.numeric(subtype)], box = T, offset = 0.1, intervals = 50, xlab = "Groups", ylab = "Brain-PAD (years)", main = "Baseline Brain-PAD by study group")) + 
  abline(0,0, lty = 2) +
  text(1,51, paste("N = ", control_n, sep = ""), cex = 0.8) +
  text(2,51, paste("N = ", rrms_n, sep = ""), cex = 0.8) +
  text(3,51, paste("N = ", spms_n, sep = ""), cex = 0.8) +
  text(4,51, paste("N = ", ppms_n, sep = ""), cex = 0.8)
}
plot.pryr
cairo_pdf("plots/MS_cohort_C3_Brain-PAD.pdf", 8,8)
plot.pryr
dev.off()
```

#### Post-hoc pairwise brain-PAD comparison of subtypes
```{r}
round(with(df.bl, TukeyHSD(aov(BrainPAD ~ subtype)))$subtype, 4)
```
### Brain-PAD by subtype descriptive statistics
```{r paged.print=FALSE}
with(df.bl, describeBy(BrainPAD, subtype, mat = T, digits = 1)) # brain-PAD by MS patient subtypes and controls
```
### Correlates of brain-PAD at baseline
#### EDSS score, an index of disability
LME accounting for fixed effects of age at baseline, gender, ICV and random effects of Cohort and scanner status.
```{r message=FALSE, warning=FALSE}
summary(lmer(BrainPAD ~ EDSSbaseline + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = df.bl))
```
When predicting brain-PAD in a LME model, the effect of EDSS at baseline beta = `r round(summary(lmer(BrainPAD ~ EDSSbaseline + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = df.bl))$coefficients["EDSSbaseline","Estimate"],2)`, 95% CI = `r round(confint(lmer(BrainPAD ~ EDSSbaseline + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = df.bl))["EDSSbaseline",],2)`, p = `r format.pval(summary(lmer(BrainPAD ~ EDSSbaseline + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = df.bl))$coefficients["EDSSbaseline","Pr(>|t|)"])`.

#### Test for interaction between subtype and EDSS on brain-PAD:
```{r}
fit.edss <- lmer(BrainPAD ~ EDSSbaseline * subtype + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "control"))
round(as.matrix(anova(fit.edss)["EDSSbaseline:subtype",]),3)
```
Use simple slopes from jtools to extract adjusted slopes for each subtype.
```{r paged.print=FALSE}
sim_slopes(fit.edss, pred = "EDSSbaseline", modx = "subtype", johnson_neyman = F)
```
Use interact_plot() from jtools to plot the adjusted slopes per group.
```{r message=FALSE, warning=FALSE}
interact_plot(fit.edss, pred = "EDSSbaseline", modx = "subtype", plot.points = T, interval = T, vary.lty = T, facet.modx = T, x.label = "EDSS", y.label = "Brain-PAD (years)", point.size = 1, modx.labels = c("CIS", "RRMS", "SPMS", "PPMS")) + geom_hline(yintercept = 0, lty = 2) + theme_bw()  +
  scale_fill_manual(values = ms.palette[2:5], name = "MS subtype") +
  scale_color_manual(values = ms.palette[2:5], name = "MS subtype")
ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/EDSS_brain-PAD_plot.pdf", height = 5, width = 8, useDingbats = FALSE)
```

### Age at diagnosis
LME accounting for fixed effects of age at baseline, gender, ICV and random effects of Cohort and scanner status.
Exclude CIS patients and healthy controls.
```{r}
summary(lmer(BrainPAD ~ disease_onset_age + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS")))
```
When predicting brain-PAD in a LME model, the effect of age at diagnosis at baseline beta = `r round(summary(lmer(BrainPAD ~ disease_onset_age + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS")))$coefficients["disease_onset_age","Estimate"],2)`, 95% CI = `r round(confint(lmer(BrainPAD ~ disease_onset_age + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS")))["disease_onset_age",],2)`, p = `r format.pval(summary(lmer(BrainPAD ~ disease_onset_age + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS")))$coefficients["disease_onset_age","Pr(>|t|)"])`.

#### Test for interaction between subtype and age at diagnosis on brain-PAD:
```{r}
fit.age <- lmer(BrainPAD ~ disease_onset_age * subtype + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS"))
round(as.matrix(anova(fit.age)["disease_onset_age:subtype",]),3)
```
Use simple slopes from jtools to extract adjusted slopes for each subtype.
```{r message=FALSE, paged.print=FALSE}
sim_slopes(fit.age, pred = "disease_onset_age", modx = "subtype", johnson_neyman = F)
```
```{r message=FALSE, warning=FALSE}
interact_plot(fit.age, pred = "disease_onset_age", modx = "subtype", plot.points = T, interval = T, vary.lty = T, facet.modx = T, x.label = "Age at clincial diagnosis (years)", y.label = "Brain-PAD (years)", point.size = 1, modx.labels = c("RRMS", "SPMS", "PPMS")) + geom_hline(yintercept = 0, lty = 2) +  theme_bw()  +
  scale_fill_manual(values = ms.palette[3:5], name = "MS subtype") +
  scale_color_manual(values = ms.palette[3:5], name = "MS subtype")
ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/diagnosis_age_brain-PAD_plot.pdf", height = 5, width = 8, useDingbats = FALSE)
```

### Time since diagnosis
LME accounting for fixed effects of age at baseline, gender, ICV and random effects of Cohort and scanner status.
Exclude controls, CIS patients and anyone with a time since diagnosis = 0.
```{r}
summary(lmer(BrainPAD ~ disease_duration_at_baseline_scan1 + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "CIS" & df.bl$disease_duration > 0)))
```
When predicting brain-PAD in a LME model, the effect of time since diagnosis at baseline beta = `r round(summary(lmer(BrainPAD ~ disease_duration_at_baseline_scan1 + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = df.bl))$coefficients["disease_duration_at_baseline_scan1","Estimate"],2)`, 95% CI = `r round(confint(lmer(BrainPAD ~ disease_duration_at_baseline_scan1 + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = df.bl))["disease_duration_at_baseline_scan1",],2)`, p = `r format.pval(summary(lmer(BrainPAD ~ disease_duration_at_baseline_scan1 + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = df.bl))$coefficients["disease_duration_at_baseline_scan1","Pr(>|t|)"])`.

#### Test for interaction between subtype and time since diagnosis on brain-PAD
```{r}
fit.time <- lmer(BrainPAD ~ disease_duration_at_baseline_scan1 * subtype + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS" & df.bl$disease_duration > 0))
round(as.matrix(anova(fit.time)["disease_duration_at_baseline_scan1:subtype",]),3)
```
Use simple slopes from jtools to extract adjusted slopes for each subtype.
```{r paged.print=FALSE}
sim_slopes(fit.time, pred = "disease_duration_at_baseline_scan1", modx = "subtype", johnson_neyman = F)
```
Plot
```{r message=FALSE, warning=FALSE}
interact_plot(fit.time, pred = "disease_duration_at_baseline_scan1", modx = "subtype", plot.points = T, interval = T, vary.lty = T, facet.modx = T, x.label = "Time since clinical diagnosis (years)", y.label = "Brain-PAD (years)", point.size = 1, modx.labels = c("RRMS", "SPMS", "PPMS")) + geom_hline(yintercept = 0, lty = 2) + theme_bw()  +
  scale_fill_manual(values = ms.palette[3:5], name = "MS subtype") +
  scale_color_manual(values = ms.palette[3:5], name = "MS subtype")
ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/diagnosis_years_brain-PAD_plot.pdf", height = 5, width = 8, useDingbats = FALSE)
```

## EDSS progression survival analysis
Based on Arman Eshaghi's code used in [Eshaghi et al., 2018 Annals of Neurology.](https://onlinelibrary.wiley.com/doi/full/10.1002/ana.25145)
```{r}
# function for characterising EDSS progression, based on different rates of change and different baseline EDSS values
is_sustained_progression <- function(edssAtStart, change){
  sustainedProgression <- FALSE
  #if start of edss is 0, 1.5 increase is considered sustained progression
  if ((edssAtStart < 1) & (change >= 1.5)) {
    sustainedProgression <- TRUE
  }
  #if start of edss is 6 or above, 0.5 increase is considered sustained progression
  else if ((edssAtStart >= 6) & (change >= 0.5 )) {
    sustainedProgression <- TRUE
  }
  #if start of edss is more than zero but less than 6, sustained progression is by 1 increase in edss
  else if ((edssAtStart >= 1 ) & (edssAtStart < 6 )  & (change >= 1  )) {
    sustainedProgression <- TRUE
  }
  return(sustainedProgression)
}

## determine change in EDSS from baseline to last follow-up  
## select latest EDSS per subject in subjects with 2 or more assessments
y1 <- df %>%
  filter(!subtype == "control")  %>%
  filter(!is.na(EDSSbaseline)) %>%
  group_by(PatientID) %>%
  top_n(1, interval) %>%
  ungroup() %>%
  dplyr::rename(latest_EDSS = EDSSatScan) %>%
  dplyr::select(PatientID, interval, EDSSbaseline, latest_EDSS) %>%
  filter(!is.na(latest_EDSS)) %>%
  mutate(EDSSchange = latest_EDSS - EDSSbaseline)

y1$EDSS_progression <-  mapply(is_sustained_progression, y1$EDSSbaseline, y1$EDSSchange) # apply Arman's function
## get baseline brain-PAD and brain volumetric measures
y2 <- df %>%
  filter(!subtype == "control")  %>%
  filter(interval == 0) %>%
  filter(!is.na(EDSSbaseline)) %>%
  dplyr::rename(BrainPAD_baseline = BrainPAD) %>%
  dplyr::rename(GM_vol_baseline = GM_vol) %>%
  dplyr::rename(WM_vol_baseline = WM_vol) %>%
  dplyr::select(-one_of('interval'))

y3 <- right_join(y1, y2, by = c("PatientID")) %>%
  filter(!is.na(latest_EDSS))
```
#### Numbers of EDSS progressors
The number of MS patients with >= 2 EDSS scores was `r y3 %>% tally()`.
```{r}
table(y3$EDSS_progression) # calculate proportion of patients who progress
round(prop.table(table(y3$EDSS_progression)),3) # calculate percentage of patients who progress
```
#### Run survival analysis
```{r}
# creating new response function
S <- Surv(time = y3$interval, event = y3$EDSS_progression)
# Brain-PAD, age, sex model
m1 <- coxph(S ~ BrainPAD_baseline + age_at_baseline_scan1 + gender, data = y3)
summary(m1)
cox.zph(m1)
```
The hazard ratio for brain-PAD on time-to-disease-progression was HR (95% CI) = `r round(summary(m1)$conf.int['BrainPAD_baseline',-2],3)`. That means for every additional +1 year of brain-PAD there is a `r round(summary(m1)$conf.int['BrainPAD_baseline',"exp(coef)"],3)`% increase in the likelihood of EDSS progression. Extrapolated over 5 years of brain-PAD, there is a `r round(summary(m1)$conf.int['BrainPAD_baseline',"exp(coef)"]^5,3)` increase in the likelihood of EDSS progression.

#### Time-to-EDSS progression Kaplan-Meier plots
```{r}
# Brain-PAD median split
km.plot.df <- y3 %>% mutate(split_BrainPAD = ntile(BrainPAD_baseline, 2))
```
Based on a median split of brain-PAD. The median value = `r round(median(km.plot.df$BrainPAD_baseline),2)` years.
```{r}
# Run survplot on survival object
S <- Surv(time = km.plot.df$interval, event = km.plot.df$EDSS_progression) # response function
survplot <- ggsurvplot(survfit(S ~ split_BrainPAD, data = km.plot.df), surv.plot.height = 0.9, ggtheme = theme_survminer(), risk.table = T, cumcensor = F, conf.int = F, palette = c("blue", "red"), censor = F , legend.labs = c("Brain-PAD < median", "Brain-PAD > median"))
survplot
ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/KM_brain-PAD_plot.pdf", height = 8, width = 8, print(survplot), useDingbats = FALSE)
```

## Longitudinal brain-age analysis
The total number of people with two or more scans was n = `r dim(subset(df.bl, df.bl$NoScans >= 2))[1]`. 

#### determine change in brain-PAD from baseline to last follow-up
```{r}
## select latest brain-PAD per subject in subjects with 2 or more assessments
z1 <- df %>%
  filter(NoScans >= 2)  %>%
  group_by(PatientID) %>%
  top_n(1, interval) %>%
  ungroup() %>%
  dplyr::rename(latest_BrainPAD = BrainPAD) %>%
  dplyr::select(PatientID, interval, latest_BrainPAD) 

## baseline brain-PAD
z2 <- df %>%
  filter(NoScans >= 2)  %>%
  filter(interval == 0) %>%
  filter(!is.na(BrainPAD)) %>%
  dplyr::rename(BrainPAD_baseline = BrainPAD) %>%
  dplyr::rename(GM_vol_baseline = GM_vol) %>%
  dplyr::rename(WM_vol_baseline = WM_vol) %>%
  dplyr::select(-one_of('interval'))

## calculate change in brain-PAD between baseline and latest brain-PAD
z3 <- right_join(z1, z2, by = c("PatientID")) %>%
  mutate(BrainPAD_change = latest_BrainPAD - BrainPAD_baseline)
```
#### Mean annualised rates of change in brain-PAD per group
```{r paged.print=FALSE}
describeBy(z3$BrainPAD_change/z3$interval, z3$subtype, mat = T, digits = 2)
```
#### determine change in EDSS from baseline to last follow-up
```{r}
## select latest EDSS per subject in subjects with 2 or more assessments
a1 <- df %>%
  filter(NoScans >= 2)  %>%
  group_by(PatientID) %>%
  top_n(1, interval) %>%
  ungroup() %>%
  dplyr::rename(latest_EDSSatScan = EDSSatScan) %>%
  dplyr::select(PatientID, interval, latest_EDSSatScan) 

## baseline EDSS
a2 <- df %>%
  filter(NoScans >= 2)  %>%
  filter(interval == 0) %>%
  filter(!is.na(EDSSatScan)) %>%
  dplyr::rename(EDSSatScan_baseline = EDSSatScan) %>%
  dplyr::rename(BrainPAD_baseline = BrainPAD) %>%
  dplyr::rename(GM_vol_baseline = GM_vol) %>%
  dplyr::rename(WM_vol_baseline = WM_vol) %>%
  dplyr::select(-one_of('interval'))

## calculate change in brain-PAD between baseline and latest brain-PAD
a3 <- right_join(a1, a2, by = c("PatientID")) %>%
  mutate(EDSS_change = latest_EDSSatScan - EDSSatScan_baseline)
```
#### Mean annualised rates of change in EDSS per group
```{r paged.print=FALSE}
with(subset(a3, a3$control0rest1 != "control"), describeBy(EDSS_change/interval, subtype, mat = F, digits = 2))
```
#### Relationship between annualised EDSS change and brain-PAD change

```{r message=FALSE}
delta.df <- join(a3, z3)
cor.test(delta.df$EDSS_change, delta.df$BrainPAD_change, method = "pearson")
```
Historgams of baseline EDSS scores and EDSS changes
```{r warning=FALSE}
plot1 <- ggplot(df.bl, aes(x = EDSSbaseline)) + geom_histogram(binwidth = 1, fill = "dodgerblue", colour = "black") + xlab("EDSS at baseline") + theme_bw()
plot2 <- ggplot(delta.df, aes(x = EDSS_change)) + geom_histogram(binwidth = 1, fill = "dodgerblue", colour = "black") + xlab("EDSS change") + theme_bw()
plot_grid(plot1, plot2, labels = c("A", "B"))
```

#### Interaction between subtype and EDSS change
```{r}
fit.change <- lm(BrainPAD_change ~ EDSS_change * subtype, data = delta.df)
anova(fit.change)
```

Use jtools package to get slopes from the model, per subtype.
```{r paged.print=FALSE}
sim_slopes(fit.change, pred = "EDSS_change", modx = "subtype", johnson_neyman = F, digits = 4)
```
Plot
```{r message=FALSE, warning=FALSE}
interact_plot(fit.change, pred = "EDSS_change", modx = "subtype", plot.points = T, interval = T, facet.modx = T, x.label = "EDSS annualised change", y.label = "Brain-PAD annualised change", modx.labels = c("CIS", "RRMS", "SPMS", "PPMS")) + geom_hline(yintercept = 0, lty = 2) + theme_bw()  +
  scale_fill_manual(values = ms.palette[-1], name = "MS subtype") +
  scale_color_manual(values = ms.palette[-1], name = "MS subtype")
ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/change_EDSS_brain-PAD_plot.pdf", height = 5, width = 8, useDingbats = FALSE)
```
Correlate baseline brain-PAD with the number of follow-up scans completed in the n=`r with(subset(df.bl, df.bl$NoScans > 1 & df.bl$subtype == "SPMS"), table(subtype))["SPMS"]` with >1 scan.
```{r}
with(subset(df.bl, df.bl$NoScans > 1 & df.bl$subtype == "SPMS"), cor.test(BrainPAD, NoScans, method = "spearman"))
```

#### Longitudinal brain-predicted age trajectories
Interaction between group and time
```{r}
## conditional growth model - random effects of participant, cohort and scanner status
model_int.group <- lmer(BrainPAD ~ control0rest1 * interval + gender + age_at_baseline_scan1 + ICV + (interval| PatientID) + (1|Cohort/scanner_status), data = df) 
round(anova(model_int.group)["control0rest1:interval",],3)
```
#### Brain-PAD change EMMs, using annualised difference between baseline and final follow-up.
Generate EMMs for all groups and MS subtypes. LME adjusting for age, gender, ICV, cohort and scanner status.
```{r}
fit <- (lmer(BrainPAD_change ~ subtype + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = z3))
emmeans(object = fit, specs = c("subtype"))
```
Generate EMMs for HCs and MS/CIS combined. LME adjusting for age, gender, ICV, cohort and scanner status.
```{r message=FALSE, warning=FALSE}
fit <- (lmer(BrainPAD_change ~ control0rest1 + age_at_baseline_scan1 + gender + ICV + (1|Cohort/scanner_status), data = z3))
emmeans(object = fit, specs = c("control0rest1"))
```
#### Slopes per group
Controls
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
control.slope <- summary(lmer(BrainPAD ~ interval + gender + age_at_baseline_scan1 + ICV + (interval| PatientID) + (1|Cohort/scanner_status), data = subset(df, df$control0rest1 == "control")))$coef["interval","Estimate"]
control.intercept <- summary(lmer(BrainPAD ~ interval + gender + age_at_baseline_scan1 + ICV + (interval| PatientID) + (1|Cohort/scanner_status), data = subset(df, df$control0rest1 == "control")))$coef["(Intercept)","Estimate"]
```
MS patients
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
ms.slope <- summary(lmer(BrainPAD ~ interval + gender + age_at_baseline_scan1 + ICV + (interval| PatientID) + (1|Cohort/scanner_status), data = subset(df, df$control0rest1 == "MS")))$coef["interval","Estimate"]
ms.intercept <- summary(lmer(BrainPAD ~ interval + gender + age_at_baseline_scan1 + ICV + (interval| PatientID) + (1|Cohort/scanner_status), data = subset(df, df$control0rest1 == "MS")))$coef["(Intercept)","Estimate"]
```

#### Longitudinal brain-PAD by interval plots
```{r}
ggplot(data = df, aes(x = interval, y = BrainPAD, fill = subtype)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_line(aes(group = PatientID, colour = subtype), alpha = 0.3, linetype = 1, size = 0.25) +
  # geom_point(aes(fill = subtype), pch = 21, size = 0.5, alpha = 0.2) +
  geom_abline(slope = control.slope, intercept = control.intercept, col = "blue") +
  geom_abline(slope = ms.slope, intercept = ms.intercept, col = "red") +
  labs(x = "Time (years)", y = "Brain-predicted age difference (years)") +
  scale_fill_manual(values = ms.palette) +
  scale_color_manual(values = ms.palette) +
  theme_bw() + theme(legend.position = "none")

ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/longitudinal_brain-PAD_time_plot.pdf", height = 6, width = 6, useDingbats = FALSE)
```


```{r}
ggplot(data = df, aes(x = age_at_scan, y = BrainPAD, fill = subtype)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_line(aes(group = PatientID, colour = subtype), alpha = 0.5, linetype = 1, size = 0.25) +
  labs(x = "Time (years)", y = "Brain-predicted age difference (years)") +
  scale_fill_manual(values = ms.palette) +
  scale_color_manual(values = ms.palette) +
  facet_wrap(~ subtype, scales = "free_x") +
  theme_bw() + theme(legend.position = "none")
```

### Hierarchical partitioning of brain-PAD
```{r paged.print=FALSE}
a <- summary(lm(BrainPAD ~ age_at_baseline_scan1 + gender + gm_vol_ratio_icv + wm_vol_ratio_icv + csf_vol_ratio_icv + Cohort, data = df.bl))
print(a)
h.p <- hier.part(y = df.bl$BrainPAD, xcan = df.bl[c("age_at_baseline_scan1","gender","gm_vol_ratio_icv", "wm_vol_ratio_icv", "csf_vol_ratio_icv", "Cohort")], gof = "Rsqu")
print(h.p)
sum(h.p$IJ["I"]) - a$adj.r.squared
sum(h.p$IJ["I"])
sum(h.p$IJ["J"])
round(h.p$IJ, 3)
round(h.p$I.perc, 1)
```
