---
title: "MS brain age UCL MAGNIMS analysis"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
  word_document: default
---
James Cole - October 2019. Built with R version `r getRversion()`  
This is Notebook contains the final brain age analysis of MS patient data and controls from the UCL cohort, the MAGNIMS consortium and the Imperial College London PET study (n=25). The analysis uses brain-predicted age difference (brain-PAD) to look at brain ageing in the context of MS. The brain-PAD values were generated in PRONTO, using an independent healthy (n=2001) training dataset, and the values were corrected for proportional bias using the intercept and slope of the age by brain-predicted age regression in the training dataset.
  
# Set-up
```{r setup, include=FALSE, cache = FALSE}
## libraries
Packages <- c("car", "corrplot", "cowplot", "data.table", "dplyr", "emmeans", "ggstance", "ggplot2", "grDevices", "gridExtra", "hier.part", "jtools", "knitr", "lmerTest", "MASS", "metafor", "plotrix", "plyr", "pryr", "psych", "qwraps2", "RColorBrewer", "scales", "stringr", "survival", "survminer", "tidyr")
lapply(Packages, library, character.only = TRUE)

## setting working directory
opts_knit$set(root.dir = "/Users/jcole/Work/Brain ageing/Collaborations/MS", paged.print = FALSE)
```

## Clear workspace, set colour palette
```{r message=FALSE, warning=FALSE}
rm(list = ls()) ## clear workspace
ms.palette <- c("darkgreen", "darkorange", "red", "blue", "purple") # define MS colour scheme for groups
sessionInfo()
```
## Get data from CSV and define longitudinal data.frames
```{r}
df <- read.csv("MS_brain_age_final_long.csv")
df$subtype <- factor(df$subtype, levels = c("control", "CIS", "RRMS", "SPMS", "PPMS")) # reorder subtype factor to put controls first
df$Cohort <- recode(df$Cohort, JR1 = "Imperial", C0 = "UCL0", C1 = "UCL1", C2 = "UCL2", C3 = "UCL3", C4 = "UCL4" ,C5 = "UCL5", C6 = "UCL6", C7 = "UCL7", A = "Amsterdam", B = "Barcelona", G = "Graz", M = "Milan", R = "Rome", S = "Siena") 
```

## Exclude participants with errors in the database & correct time since diagnosis errors
```{r message=FALSE, warning=FALSE}
tmp <- df[df$Cohort == 'Amsterdam',] %>% group_by(PatientID) %>% dplyr::summarize(sd = sd(age_at_scan)) %>% arrange(desc(sd)) %>% filter(sd > 2)
excluded_IDs <- sort(tmp$PatientID)
# excluded_IDs <- unique(df[which(df$age_at_scan < df$age_at_baseline_scan1),1])
excluded_scans <- (df %>% filter(str_detect(PatientID, paste(excluded_IDs, collapse = "|"))))["ScanName"]
df <- df %>%
  filter(!str_detect(PatientID, paste(excluded_IDs, collapse = "|")))
rm(tmp)
```
There were `r length(excluded_IDs)` subjects with `r dim(excluded_scans)[1]` scans excluded in total. Data entry errors in original spreadsheet; age at baseline not consistent within subject.

## Load data for lesion filling analysis
```{r}
## read in data
lesion_df <- read.csv(file = '~/Work/Brain ageing/Collaborations/MS/MAGNIMS/MAGNIMS20171224_final.csv')
lesion_df$subtype <- factor(lesion_df$subtype, levels = c("control", "CIS", "RRMS", "SPMS", "PPMS"))
## create ICV ratio variables
lesion_df <- lesion_df %>%
  mutate(ICV = GMV + WMV + CSFV) %>%
  mutate(gm_icv_ratio = GMV/ICV) %>%
  mutate(wm_icv_ratio = WMV/ICV)
```

## Recode scanner status variable to field strength
Pre-post 2004 data only available for one site.
```{r}
table(df$scanner_status)
df$field_strength <- recode(df$scanner_status, `1.5T_pre_2004` = "1.5T", `1.5T_post_2004` = "1.5T")
table(df$field_strength)
```


## Generate baseline only data.frame and show data frame structure
```{r}
df.bl <- df[(df$age_at_baseline_scan1 == df$age_at_scan),] # define baseline data.frame
str(df)
```

# Basic stats
## Total number of subjects, and by group
The total number of subjects included was n = `r dim(df.bl)[1]`

The total number of MS patients (including CIS) was  n = `r table(df.bl$control0rest1)[2]` and healthy controls was n = `r table(df.bl$control0rest1)[1]`

## Number of scans in total and per group
Total number of scans = `r sum(df.bl$NoScans)`

## Number of people with 2 or more scans
```{r}
df.bl %>%
  filter(NoScans >= 2) %>%
  group_by(control0rest1) %>%
  tally()
```
## Number of people with 3 or more scans
```{r}
df.bl %>%
  filter(NoScans >= 3) %>%
  group_by(control0rest1) %>%
  tally()
```
## Generate demographics table using qwarps2
```{r message=FALSE, warning=FALSE}
options(qwraps2_markup = "markdown", digits = 2)

table1 <-
  list("N" =
         list("Control" = ~ qwraps2::n_perc0(control0rest1 == "control", na_rm = T),
              "MS"  = ~ qwraps2::n_perc0(control0rest1 == "MS", na_rm = T)),
     "Gender" =
       list("Female" = ~ qwraps2::n_perc0(gender == "female", show_symbol = T),
            "Male"  = ~ qwraps2::n_perc0(gender == "male", show_symbol = T)),
        "Number of scans" =
       list("min" = ~ min(NoScans),
            "max" = ~ max(NoScans),
            "mean (sd)" = ~ qwraps2::mean_sd(NoScans)),
       "Age at baseline scan (years)" =
       list("min" = ~ min(age_at_baseline_scan1),
            "max" = ~ max(age_at_baseline_scan1),
            "mean (sd)" = ~ qwraps2::mean_sd(age_at_baseline_scan1)),
       "Brain-predicted age at baseline scan (years)" =
       list("min" = ~ min(BrainAge),
            "max" = ~ max(BrainAge),
            "mean (sd)" = ~ qwraps2::mean_sd(BrainAge)),
       "Disease duration at baseline (years)" =
       list("min" = ~ min(disease_duration, na.rm = T),
            "max" = ~ max(disease_duration, na.rm = T),
            "mean (sd)" = ~ qwraps2::mean_sd(disease_duration, na_rm = T, show_n = "never")),
        "EDSS at baseline " =
       list("min" = ~ min(EDSSbaseline, na.rm = T),
            "max" = ~ max(EDSSbaseline, na.rm = T),
            "mean (sd)" = ~ qwraps2::mean_sd(EDSSbaseline, na_rm = T, show_n = "never")),
       "MS subtype" =
       list("CIS" = ~ qwraps2::n_perc0(subtype == "CIS", show_symbol = T),
            "RRMS"  = ~ qwraps2::n_perc0(subtype == "RRMS", show_symbol = T),
            "SPMS"  = ~ qwraps2::n_perc0(subtype == "SPMS", show_symbol = T),
            "PPMS"  = ~ qwraps2::n_perc0(subtype == "PPMS", show_symbol = T)),
       "Treatment" =
         list("Yes" = ~ qwraps2::n_perc0(DMT_YesNoNA == "YES", na_rm = T, show_symbol = T),
              "No"  = ~ qwraps2::n_perc0(DMT_YesNoNA == "NO", na_rm = T, show_symbol = T))
       )

print(summary_table(dplyr::group_by(df.bl, control0rest1), table1),
      rtitle = "Sample Chararcteristics",
      cnames = c("Controls", "MS patients"))
```
### Need to get treatment NAs using table()
```{r}
table(df.bl$DMT_YesNoNA, df.bl$control0rest1, useNA = "ifany")[,"MS"]
```

## Length of follow-up
Get length of follow-up from longitudinal database.
```{r}
options(digits = 3) ## return digits option to default
df %>%
  filter(NoScans >= 2) %>%
  group_by(PatientID) %>%
  slice(which.max(interval)) %>%
  # top_n(n = 1, wt = interval) %>%
  group_by(control0rest1) %>%
  dplyr::summarise(mean(interval), sd(interval), min(interval), max(interval))
options(digits = 7) ## return digits option to default
```

## Correlations between demographics and clinical variables
Use corrplot package.
```{r}
cor.mat <- cbind(df.bl$age_at_scan, df.bl$disease_onset_age, df.bl$disease_duration, df.bl$EDSSatScan)
colnames(cor.mat) <- c("Age at scan", "Age at diagnosis", "Time since diagnosis", "EDSS at scan")
corrplot(cor(cor.mat, use = "pairwise"), type = "upper", method = "color", addCoef.col = T, tl.col = "black", diag = F)
```

# Baseline brain-age analysis
```{r paged.print=FALSE}
describeBy(df.bl$BrainPAD, df.bl$control0rest1, mat = T, digits = 3) # brain-PAD by MS patient vs. controls
```

## Evalute potential covariates

```{r}
table(df.bl$Cohort, df.bl$field_strength)
```

```{r}
anova(lm(BrainPAD ~ poly(age_at_baseline_scan1, 2) + gender + ICV + field_strength + Cohort, data = df.bl))
```

## Hierarchical partitioning of brain-PAD
```{r paged.print=FALSE}
a <- summary(lm(BrainPAD ~ poly(age_at_baseline_scan1, 2) + gender + gm_vol_ratio_icv + wm_vol_ratio_icv + csf_vol_ratio_icv + field_strength + Cohort, data = df.bl))
print(a)
h.p <- hier.part(y = df.bl$BrainPAD, xcan = df.bl[c("age_at_baseline_scan1","gender","gm_vol_ratio_icv", "wm_vol_ratio_icv", "csf_vol_ratio_icv", "field_strength", "Cohort")], gof = "Rsqu")
round(h.p$IJ, 3)
round(h.p$I.perc, 1)
```

Results suggest that age, age^2, gender, Cohort and field strength are appropriate covariates.
# Predict brain-PAD based on group
Function to run a linear mixed effect (LME) model adjusting for: fixed effects of age, age^2, gender and field strength; random effects of cohort.
```{r}
run_lm <- function(var, data) {
  m1 <- lmer(BrainPAD ~ data[[var]] +
               poly(age_at_scan, 2) +
               gender +
               field_strength +
               (1|Cohort),
             data = data,
             control = lmerControl(optimizer = "Nelder_Mead"))
  return(m1)
}
```

## Main effect of group (MS vs. controls)
```{r}
fit <- run_lm("control0rest1", df.bl)
summary(fit)
```
## Estimated marginal means
Generate EMMs for all MS/CIS and healthy controls.
```{r}
fit <- lmer(BrainPAD ~ control0rest1 +
               poly(age_at_scan, 2) +
               gender +
               field_strength +
               (1|Cohort),
             data = df.bl,
             control = lmerControl(optimizer = "Nelder_Mead"))
emmeans(object = fit, ~ control0rest1)
```

# Meta-analysis looking at all the separate cohorts with MS/CIS patients and controls
### Check which cohorts contain healthy controls and patients.
```{r}
table(df.bl$subtype, df.bl$Cohort)
```

### Create data.frame with summary data appropriate for meta-analysis.
```{r}
tmp0 <- df.bl %>% group_by(Cohort, control0rest1) %>% dplyr::summarise(length(control0rest1))
meta.cohorts <- as.list(tmp0[tmp0$control0rest1 == "control",1])$Cohort
tmp <- df.bl %>%
  filter(str_detect(Cohort, paste(meta.cohorts, collapse = "|")))
tmp1 <- tmp %>% group_by(Cohort, control0rest1) %>% dplyr::summarise(n = n(), Mean = mean(BrainPAD), SD  =  sd(BrainPAD))
tmp2 <- dcast(tmp1, Cohort ~ control0rest1, value.var = "n")
tmp3 <- dcast(tmp1, Cohort ~ control0rest1, value.var = "Mean")
tmp4 <- dcast(tmp1, Cohort ~ control0rest1, value.var = "SD")
names(tmp2) <- c("Cohort", "control_n", "MS_n")
names(tmp3) <- c("Cohort", "control_mean", "MS_mean")
names(tmp4) <- c("Cohort", "control_sd", "MS_sd")
tmp1x <- tmp %>% group_by(Cohort) %>% dplyr::summarise(pooled.SD  =  sd(BrainPAD))
meta.df <- join_all(list(tmp2,tmp3,tmp4, tmp1x), by = 'Cohort', type = 'left')
rm(list = ls(pattern = 'tmp*')) # remove temporary data frames
```

## Fit a random-effects meta-analysis using REML
Using the metafor package.
```{r}
meta.df <- escalc(m1i = control_mean, sd1i = pooled.SD, n1i = control_n, m2i = MS_mean, sd2i = pooled.SD, n2i = MS_n, measure = "MD", data = meta.df, digits = 2)
meta.results <- rma(yi, vi, data = meta.df, method = "REML")
print(meta.results)
confint(meta.results)
```

## Forest plot of results
```{r message=FALSE, warning=FALSE}
plot.forest %<a-% {
forest(meta.results, ilab = cbind(meta.df$MS_n, meta.df$control_n), ilab.xpos = c(-30,-23), slab = meta.df$Cohort, digits = 1, xlab = "MS vs. Healthy control group mean difference", steps = 6, col = "red", cex = 1.25, pch = 22, bg = "blue"); text(c(-40, -30, -23), 7.6, c("Cohort", "MS n", "HC n"), font = 2, cex = 1.25)
}
plot.forest
cairo_pdf("plots/forest_plot.pdf", 6,5)
plot.forest
dev.off()
```

## Linear regression analysis in cohort UCL3
Includes covariates: age, age^2, gender.
```{r}
summary(lm(BrainPAD ~ control0rest1 + poly(age_at_baseline_scan1, 2) + gender, data = subset(df.bl, df.bl$Cohort == "UCL3")))
```

# Brain-PAD by MS subtype
Using a linear mixed effect (LME) model, to compare subtypes. This analysis excluded controls. Adjusting for age, age^2, gender, field strength and cohort.
```{r}
subtypes <- run_lm("subtype", subset(df.bl, df.bl$subtype != "control"))
summary(subtypes)
```
## Brain-PAD estimated marginal means for subtypes
Generate EMMs for all MS subtypes.
```{r}
fit <- lmer(BrainPAD ~ subtype +
               poly(age_at_scan, 2) +
               gender +
               field_strength +
               (1|Cohort),
             data = subset(df.bl, df.bl$control0rest1 != "control"),
             control = lmerControl(optimizer = "Nelder_Mead"))
emmeans(object = fit, ~ subtype)
```

## Brain-PAD boxplot by MS subtype
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# calculate Ns
control_n <- with(df.bl, table(subtype))["control"]
cis_n <- with(df.bl, table(subtype))["CIS"]
rrms_n <- with(df.bl, table(subtype))["RRMS"]
spms_n <- with(df.bl, table(subtype))["SPMS"]
ppms_n <- with(df.bl, table(subtype))["PPMS"]

plot.pryr %<a-% {
with(df.bl, ehplot(BrainPAD, groups = subtype, ylim = c(-20,52), pch = 21, bg = ms.palette[as.numeric(subtype)], box = T, offset = 0.1, intervals = 50, xlab = "Groups", ylab = "Brain-PAD (years)", main = "Baseline Brain-PAD by study group")) + 
  abline(0,0, lty = 2) +
  text(1,51, paste("N = ", control_n, sep = ""), cex = 0.8) +
  text(2,51, paste("N = ", cis_n, sep = ""), cex = 0.8) +
  text(3,51, paste("N = ", rrms_n, sep = ""), cex = 0.8) +
  text(4,51, paste("N = ", spms_n, sep = ""), cex = 0.8) +
  text(5,51, paste("N = ", ppms_n, sep = ""), cex = 0.8)
}
plot.pryr
cairo_pdf("plots/MS_UCL_MAGNIMS_combined_group_Brain-PAD.pdf", 8,8)
plot.pryr
dev.off()
```

## Brain-PAD boxplot by MS subtype in cohort UCL3 only
```{r message=FALSE, warning=FALSE}
# calculate Ns
C3.bl <- df.bl %>% filter(Cohort == "UCL3") 
C3.bl$subtype <- factor(C3.bl$subtype)
control_n <- with(C3.bl, table(subtype))["control"]
rrms_n <- with(C3.bl, table(subtype))["RRMS"]
spms_n <- with(C3.bl, table(subtype))["SPMS"]
ppms_n <- with(C3.bl, table(subtype))["PPMS"]

plot.pryr %<a-% {
  with(C3.bl, ehplot(BrainPAD, groups = subtype, ylim = c(-20,52), pch = 21, bg = ms.palette[-2][as.numeric(subtype)], box = T, offset = 0.1, intervals = 50, xlab = "Groups", ylab = "Brain-PAD (years)", main = "Baseline Brain-PAD by study group")) + 
  abline(0,0, lty = 2) +
  text(1,51, paste("N = ", control_n, sep = ""), cex = 0.8) +
  text(2,51, paste("N = ", rrms_n, sep = ""), cex = 0.8) +
  text(3,51, paste("N = ", spms_n, sep = ""), cex = 0.8) +
  text(4,51, paste("N = ", ppms_n, sep = ""), cex = 0.8)
}
plot.pryr
cairo_pdf("plots/MS_cohort_C3_Brain-PAD.pdf", 8,8)
plot.pryr
dev.off()
```

## Post-hoc pairwise brain-PAD comparison of subtypes
```{r}
round(with(df.bl, TukeyHSD(aov(BrainPAD ~ subtype)))$subtype, 4)
```

## Brain-PAD by subtype descriptive statistics
```{r paged.print=FALSE}
with(df.bl, describeBy(BrainPAD, subtype, mat = T, digits = 1)) # brain-PAD by MS patient subtypes and controls
```

# Lesion filling
To establish whether using the FSL lesion filling software influences brain-predicted age values.
This analysis was conducted only in UCL patients (n = `r table(!is.na(lesion_df$filled_brain_age))["TRUE"]`).
```{r paged.print=FALSE}
with(subset(lesion_df, lesion_df$subtype != "control"), describeBy(filled_brain_age, subtype))
```
Correlation between brain-predicted age from filled and unfilled images: Pearson's r = `r round(with(subset(lesion_df, lesion_df$subtype != "control"), cor(filled_brain_age, brain_age, use = "pairwise")),3)`.
Median absolute error (MAE) between brain-predicted age from filled and unfilled images = `r with(subset(lesion_df, lesion_df$subtype != "control"), median(abs(filled_brain_age - brain_age), na.rm = T))` years.
Mean difference between brain-predicted age from filled and unfilled images = `r round(mean(lesion_df$filled_brain_age - lesion_df$brain_age, na.rm = T),3)` ± `r round(sd(lesion_df$filled_brain_age - lesion_df$brain_age, na.rm = T),3)` years.

## Lesion filled vs. unfilled scatterplot
```{r}
ggplot(data = subset(lesion_df, lesion_df$subtype != "control"), aes(x = brain_age, y = filled_brain_age)) +
  geom_abline(slope = 1) +
  geom_point(pch = 21, aes(fill = subtype), size = 2) +
  labs(x = "Unfilled brain-predicted age (years)", y = "Lesion-filled brain-predicted age (years)") +
  xlim(c(20,90)) +
  scale_fill_manual(values = ms.palette[-1]) +
  theme_bw() + theme(legend.position = c(0.9, 0.2))
ggsave("~/Work/Brain ageing/Collaborations/MS/plots/lesion_filling_brain_age_plot.pdf", width = 5, height = 5, useDingbats = FALSE)
```
## Lesion filled vs. unfilled Bland-Altman plot
```{r}
mean.diff <- mean(lesion_df$brain_age - lesion_df$filled_brain_age, na.rm = T)
sd.diff <- sd(lesion_df$brain_age - lesion_df$filled_brain_age, na.rm = T)
ggplot(data = subset(lesion_df, lesion_df$subtype != "control"), aes(x = ((brain_age + filled_brain_age)/2), y = brain_age - filled_brain_age)) +
  geom_abline(slope = 0, lty = 2) +
  geom_point(pch = 21, aes(fill = subtype), size = 2) +
  geom_hline(yintercept = mean.diff, color = "darkgoldenrod1", lwd = 1) + # mean difference line
  geom_hline(yintercept = mean.diff + 1.96*sd.diff, color = "darkgoldenrod2", lty = 2) + # upper 95% line
  geom_hline(yintercept = mean.diff - 1.96*sd.diff, color = "darkgoldenrod2", lty = 2) + # lower 95% line
  # geom_smooth(method = "lm", level = 0.95, color = "black", lwd = 0.3) +
  labs(x = "Mean of filled/unfilled brain-predicted age (years)", y = "Unfilled - Lesion-filled brain-predicted age (years)") +
  # ylim(c(-20,20)) +
  scale_fill_manual(values = ms.palette[-1]) +
  theme_bw() + theme(legend.position = c(0.1, 0.8)) +
  annotate("text", x = 25, y = mean.diff + 1.96*sd.diff + 0.5, label = "+1.96*SD", color = "darkgoldenrod2") +
  annotate("text", x = 25, y = mean.diff - 1.96*sd.diff - 0.5, label = "-1.96*SD", color = "darkgoldenrod2") +
  annotate("text", x = 27.5, y = mean.diff + 0.8, label = "Mean difference", color = "darkgoldenrod2")
ggsave("~/Work/Brain ageing/Collaborations/MS/plots/lesion_filling_brain_age_BA_plot.pdf", width = 8, height = 5, useDingbats = FALSE)
```

# Correlates of brain-PAD at baseline
## EDSS score, an index of disability
LME model accounting for fixed effects of age at baseline, age^2, gender, and random effects of Cohort and field strength.
```{r message=FALSE, warning=FALSE}
summary(run_lm("EDSSbaseline", df.bl))
```

When predicting brain-PAD in a LME model, the effect of EDSS at baseline beta = `r round(summary(run_lm("EDSSbaseline", df.bl))$coefficients[2,1],2)`, 95% CI = `r round(confint(run_lm("EDSSbaseline", df.bl), method = "Wald")[4,],2)`, p = `r format.pval(summary(run_lm("EDSSbaseline", df.bl))$coefficients[2,"Pr(>|t|)"])`.

## Test for interaction between subtype and EDSS on brain-PAD
```{r}
fit.edss <- lmer(BrainPAD ~ EDSSbaseline * subtype + poly(age_at_baseline_scan1,2) + gender + field_strength + (1|Cohort), data = subset(df.bl, df.bl$subtype != "control"))
round(as.matrix(anova(fit.edss)["EDSSbaseline:subtype",]),3)
```
Use simple slopes from jtools to extract adjusted slopes for each subtype.
Need to fit model without age^2 as poly() is incompatible with sim_slopes().
```{r paged.print=FALSE}
fit.edss2 <- lmer(BrainPAD ~ EDSSbaseline * subtype + age_at_baseline_scan1 + gender + field_strength + (1|Cohort), data = subset(df.bl, df.bl$subtype != "control"))
sim_slopes(fit.edss2, pred = "EDSSbaseline", modx = "subtype", johnson_neyman = F)
```
Use interact_plot() from jtools to plot the adjusted slopes per group.
```{r message=FALSE, warning=FALSE}
edss.plot <- interact_plot(fit.edss2, pred = "EDSSbaseline", modx = "subtype", plot.points = T, interval = T, vary.lty = T, facet.modx = T, x.label = "EDSS", y.label = "Brain-PAD (years)", point.size = 1, modx.labels = c("CIS", "RRMS", "SPMS", "PPMS")) +
  geom_hline(yintercept = 0, lty = 2) +
  theme_bw() +
  scale_fill_manual(values = ms.palette[2:5], name = "MS subtype") +
  scale_color_manual(values = ms.palette[2:5], name = "MS subtype")
# ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/EDSS_brain-PAD_plot.pdf", height = 5, width = 8, useDingbats = FALSE)
```

## Age at diagnosis
LME accounting for fixed effects of age at baseline, age^2, gender, and random effects of Cohort and field strength.
Exclude CIS patients and healthy controls.
```{r}
summary(run_lm("disease_onset_age", subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS")))
```

When predicting brain-PAD in a LME model, the effect of age at diagnosis at baseline beta = `r round(summary(run_lm("disease_onset_age", subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS")))$coefficients[2,1],2)`, 95% CI = `r round(confint(run_lm("disease_onset_age", subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS")), method = "Wald")[4,],2)`, p = `r format.pval(summary(run_lm("disease_onset_age", subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS")))$coefficients[2,"Pr(>|t|)"])`.

### Test for interaction between subtype and age at diagnosis on brain-PAD:
```{r}
fit.age <- lmer(BrainPAD ~ disease_onset_age * subtype + poly(age_at_baseline_scan1,2) + gender + field_strength + (1|Cohort), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS"))
round(as.matrix(anova(fit.age)["disease_onset_age:subtype",]),3)
```
Use simple slopes from jtools to extract adjusted slopes for each subtype.
```{r message=FALSE, paged.print=FALSE}
fit.age2 <- lmer(BrainPAD ~ disease_onset_age * subtype + age_at_baseline_scan1 + gender + field_strength + (1|Cohort), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS"))
sim_slopes(fit.age2, pred = "disease_onset_age", modx = "subtype", johnson_neyman = F)
```

```{r message=FALSE, warning=FALSE}
age.plot <- interact_plot(fit.age2, pred = "disease_onset_age", modx = "subtype", plot.points = T, interval = T, vary.lty = T, facet.modx = T, x.label = "Age at clincial diagnosis (years)", y.label = "Brain-PAD (years)", point.size = 1, modx.labels = c("RRMS", "SPMS", "PPMS")) + geom_hline(yintercept = 0, lty = 2) +  theme_bw()  +
  scale_fill_manual(values = ms.palette[3:5], name = "MS subtype") +
  scale_color_manual(values = ms.palette[3:5], name = "MS subtype")
# ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/diagnosis_age_brain-PAD_plot.pdf", height = 5, width = 8, useDingbats = FALSE)
```

## Time since diagnosis
LME accounting for fixed effects of age at baseline, age^2 gender, and random effects of Cohort and field strength.
Exclude controls, CIS patients and anyone with a time since diagnosis = 0.
```{r}
summary(run_lm("disease_duration_at_baseline_scan1", subset(df.bl, df.bl$subtype != "CIS" & df.bl$disease_duration > 0)))
```

When predicting brain-PAD in a LME model, the effect of time since diagnosis at baseline beta = `r round(summary(run_lm("disease_duration_at_baseline_scan1", subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS"  & df.bl$disease_duration > 0)))$coefficients[2,1],2)`, 95% CI = `r round(confint(run_lm("disease_duration_at_baseline_scan1", subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS"  & df.bl$disease_duration > 0)), method = "Wald")[4,],2)`, p = `r format.pval(summary(run_lm("disease_duration_at_baseline_scan1", subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS"  & df.bl$disease_duration > 0)))$coefficients[2,"Pr(>|t|)"])`.

### Test for interaction between subtype and time since diagnosis on brain-PAD
```{r}
fit.time <- lmer(BrainPAD ~ disease_duration_at_baseline_scan1 * subtype + poly(age_at_baseline_scan1,2) + gender + field_strength + (1|Cohort), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS" & df.bl$disease_duration > 0))
round(as.matrix(anova(fit.time)["disease_duration_at_baseline_scan1:subtype",]),3)
```
Use simple slopes from jtools to extract adjusted slopes for each subtype.
```{r paged.print=FALSE}
fit.time2 <- lmer(BrainPAD ~ disease_duration_at_baseline_scan1 * subtype + age_at_baseline_scan1 + gender + field_strength + (1|Cohort), data = subset(df.bl, df.bl$subtype != "control" & df.bl$subtype != "CIS" & df.bl$disease_duration > 0))
sim_slopes(fit.time2, pred = "disease_duration_at_baseline_scan1", modx = "subtype", johnson_neyman = F)
```

```{r message=FALSE, warning=FALSE}
time.plot <- interact_plot(fit.time2, pred = "disease_duration_at_baseline_scan1", modx = "subtype", plot.points = T, interval = T, vary.lty = T, facet.modx = T, x.label = "Time since clinical diagnosis (years)", y.label = "Brain-PAD (years)", point.size = 1, modx.labels = c("RRMS", "SPMS", "PPMS")) + geom_hline(yintercept = 0, lty = 2) + theme_bw()  +
  scale_fill_manual(values = ms.palette[3:5], name = "MS subtype") +
  scale_color_manual(values = ms.palette[3:5], name = "MS subtype")
# ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/diagnosis_years_brain-PAD_plot.pdf", height = 5, width = 8, useDingbats = FALSE)
```

## Arrange EDSS, age at diagnosis and time since diagnosis plots
Use cowplot package.
```{r}
plot_grid(edss.plot, age.plot, time.plot, labels = "AUTO", ncol = 1)
ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/clinical_brain-PAD_plots.pdf", height = 15, width = 8, useDingbats = FALSE)
```

# EDSS progression survival analysis
Based on Arman Eshaghi's code used in [Eshaghi et al., 2018 Annals of Neurology.](https://onlinelibrary.wiley.com/doi/full/10.1002/ana.25145)
Function for characterising EDSS progression, based on different rates of change and different baseline EDSS values
```{r}
is_sustained_progression <- function(edssAtStart, change){
  sustainedProgression <- FALSE
  #if start of edss is 0, 1.5 increase is considered sustained progression
  if ((edssAtStart < 1) & (change >= 1.5)) {
    sustainedProgression <- TRUE
  }
  #if start of edss is 6 or above, 0.5 increase is considered sustained progression
  else if ((edssAtStart >= 6) & (change >= 0.5 )) {
    sustainedProgression <- TRUE
  }
  #if start of edss is more than zero but less than 6, sustained progression is by 1 increase in edss
  else if ((edssAtStart >= 1 ) & (edssAtStart < 6 )  & (change >= 1  )) {
    sustainedProgression <- TRUE
  }
  return(sustainedProgression)
}
```

### determine change in EDSS from baseline to last follow-up
Select latest EDSS per subject in subjects with 2 or more assessments
```{r}
y1 <- df %>%
  filter(!subtype == "control")  %>%
  filter(!is.na(EDSSbaseline)) %>%
  group_by(PatientID) %>%
  top_n(1, interval) %>%
  ungroup() %>%
  dplyr::rename(latest_EDSS = EDSSatScan) %>%
  dplyr::select(PatientID, interval, EDSSbaseline, latest_EDSS) %>%
  filter(!is.na(latest_EDSS)) %>%
  mutate(EDSSchange = latest_EDSS - EDSSbaseline)

y1$EDSS_progression <-  mapply(is_sustained_progression, y1$EDSSbaseline, y1$EDSSchange) # apply Arman's function
## get baseline brain-PAD and brain volumetric measures
y2 <- df %>%
  filter(!subtype == "control")  %>%
  filter(interval == 0) %>%
  filter(!is.na(EDSSbaseline)) %>%
  dplyr::rename(BrainPAD_baseline = BrainPAD) %>%
  dplyr::rename(GM_vol_baseline = GM_vol) %>%
  dplyr::rename(WM_vol_baseline = WM_vol) %>%
  dplyr::rename(CSF_vol_baseline = CSF_vol) %>%
  dplyr::rename(WBV_baseline = WBV) %>%
  dplyr::rename(ICV_baseline = ICV) %>%
  dplyr::select(-one_of('interval'))

y3 <- right_join(y1, y2, by = c("PatientID")) %>%
  filter(!is.na(latest_EDSS))
```

### Numbers of EDSS progressors
The number of MS patients with >= 2 EDSS scores was `r y3 %>% tally()`.
```{r}
table(y3$EDSS_progression) # calculate proportion of patients who progress
round(prop.table(table(y3$EDSS_progression)),3) # calculate percentage of patients who progress
```
## Run survival analysis
```{r}
# creating new response function
S <- Surv(time = y3$interval, event = y3$EDSS_progression)
# Brain-PAD, age, sex model
surv.model <- coxph(S ~ BrainPAD_baseline + age_at_baseline_scan1 + gender, data = y3)
summary(surv.model)
# Check the assumptions of proportional hazards are met
cox.zph(surv.model)
```
Check that the brain-PAD line is horizontal.
```{r}
plot(cox.zph(surv.model)[1])
```
The hazard ratio for brain-PAD on time-to-disease-progression was HR (95% CI) = `r round(summary(surv.model)$conf.int['BrainPAD_baseline',-2],3)`. That means for every additional +1 year of brain-PAD there is a `r round(summary(surv.model)$conf.int['BrainPAD_baseline',"exp(coef)"],3)`% increase in the likelihood of EDSS progression. Extrapolated over 5 years of brain-PAD, there is a `r round(summary(surv.model)$conf.int['BrainPAD_baseline',"exp(coef)"]^5,3)` increase in the likelihood of EDSS progression.

## Time-to-EDSS progression Kaplan-Meier plots
```{r}
# Brain-PAD median split
km.plot.df <- y3 %>% mutate(split_BrainPAD = ntile(BrainPAD_baseline, 2))
```

Based on a median split of brain-PAD. The median value = `r round(median(km.plot.df$BrainPAD_baseline),2)` years.
```{r}
# Run survplot on survival object
S <- Surv(time = km.plot.df$interval, event = km.plot.df$EDSS_progression) # response function
survplot <- ggsurvplot(survfit(S ~ split_BrainPAD, data = km.plot.df), surv.plot.height = 0.9, ggtheme = theme_survminer(), risk.table = T, cumcensor = F, conf.int = F, palette = c("blue", "red"), censor = F , legend.labs = c("Brain-PAD < median", "Brain-PAD > median"))
survplot
ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/KM_brain-PAD_plot.pdf", height = 8, width = 8, print(survplot), useDingbats = FALSE)
```

# Longitudinal brain-age analysis
The total number of people with two or more scans was n = `r dim(subset(df.bl, df.bl$NoScans >= 2))[1]`. This included n = `r dim(subset(df.bl, df.bl$subtype != "control" & df.bl$NoScans >= 2))[1]` MS patients and n = `r dim(subset(df.bl, df.bl$subtype == "control" & df.bl$NoScans >= 2))[1]` controls.

With 3 or more scans, there were n = `r dim(subset(df.bl, df.bl$subtype != "control" & df.bl$NoScans >= 3))[1]` MS patients and n = `r dim(subset(df.bl, df.bl$subtype == "control" & df.bl$NoScans >= 3))[1]` controls.

## determine change in brain-PAD from baseline to last follow-up
```{r}
## select latest brain-PAD per subject in subjects with 2 or more assessments
z1 <- df %>%
  filter(NoScans >= 2)  %>%
  group_by(PatientID) %>%
  top_n(1, interval) %>%
  ungroup() %>%
  dplyr::rename(latest_BrainPAD = BrainPAD) %>%
  dplyr::select(PatientID, interval, latest_BrainPAD) 

## baseline brain-PAD
z2 <- df %>%
  filter(NoScans >= 2)  %>%
  filter(interval == 0) %>%
  filter(!is.na(BrainPAD)) %>%
  dplyr::rename(BrainPAD_baseline = BrainPAD) %>%
  dplyr::rename(GM_vol_baseline = GM_vol) %>%
  dplyr::rename(WM_vol_baseline = WM_vol) %>%
  dplyr::select(-one_of('interval'))

## calculate change in brain-PAD between baseline and latest brain-PAD
z3 <- right_join(z1, z2, by = c("PatientID")) %>%
  mutate(BrainPAD_change = latest_BrainPAD - BrainPAD_baseline)
```

## Mean annualised rates of change in brain-PAD per group
```{r paged.print=FALSE}
describeBy(z3$BrainPAD_change/z3$interval, z3$subtype, mat = T, digits = 2)
```

## determine change in EDSS from baseline to last follow-up
```{r}
## select latest EDSS per subject in subjects with 2 or more assessments
a1 <- df %>%
  filter(NoScans >= 2)  %>%
  group_by(PatientID) %>%
  top_n(1, interval) %>%
  ungroup() %>%
  dplyr::rename(latest_EDSSatScan = EDSSatScan) %>%
  dplyr::select(PatientID, interval, latest_EDSSatScan) 

## baseline EDSS
a2 <- df %>%
  filter(NoScans >= 2)  %>%
  filter(interval == 0) %>%
  filter(!is.na(EDSSatScan)) %>%
  dplyr::rename(EDSSatScan_baseline = EDSSatScan) %>%
  dplyr::rename(BrainPAD_baseline = BrainPAD) %>%
  dplyr::rename(GM_vol_baseline = GM_vol) %>%
  dplyr::rename(WM_vol_baseline = WM_vol) %>%
  dplyr::select(-one_of('interval'))

## calculate change in brain-PAD between baseline and latest brain-PAD
a3 <- right_join(a1, a2, by = c("PatientID")) %>%
  mutate(EDSS_change = latest_EDSSatScan - EDSSatScan_baseline)
```

## Mean annualised rates of change in EDSS per group
```{r paged.print=FALSE}
with(subset(a3, a3$control0rest1 != "control"), describeBy(EDSS_change/interval, subtype, mat = F, digits = 2))
```

## Relationship between annualised EDSS change and brain-PAD change
The total number of patients with two or more scans was n = `r dim(subset(df.bl, df.bl$subtype != "control" & df.bl$NoScans >= 2))[1]`. 
```{r message=FALSE}
delta.df <- join(a3, z3)
cor.test(delta.df$EDSS_change, delta.df$BrainPAD_change, method = "pearson")
```
### Historgams of baseline EDSS scores and EDSS changes
```{r warning=FALSE}
plot1 <- ggplot(df.bl, aes(x = EDSSbaseline)) + geom_histogram(binwidth = 1, fill = "dodgerblue", colour = "black") + xlab("EDSS at baseline") + theme_bw()
plot2 <- ggplot(delta.df, aes(x = EDSS_change)) + geom_histogram(binwidth = 1, fill = "dodgerblue", colour = "black") + xlab("EDSS change") + theme_bw()
plot_grid(plot1, plot2, labels = c("A", "B"))
```

## Interaction between subtype and EDSS change
```{r}
fit.change <- lm(BrainPAD_change ~ EDSS_change * subtype, data = delta.df)
anova(fit.change)
```

Use jtools package to get slopes from the model, per subtype.
```{r paged.print=FALSE}
sim_slopes(fit.change, pred = "EDSS_change", modx = "subtype", johnson_neyman = F, digits = 4)
```

```{r message=FALSE, warning=FALSE}
interact_plot(fit.change, pred = "EDSS_change", modx = "subtype", plot.points = T, interval = T, facet.modx = T, x.label = "EDSS annualised change", y.label = "Brain-PAD annualised change", modx.labels = c("CIS", "RRMS", "SPMS", "PPMS")) + geom_hline(yintercept = 0, lty = 2) + theme_bw()  +
  scale_fill_manual(values = ms.palette[-1], name = "MS subtype") +
  scale_color_manual(values = ms.palette[-1], name = "MS subtype")
ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/change_EDSS_brain-PAD_plot.pdf", height = 5, width = 8, useDingbats = FALSE)
```
Correlate baseline brain-PAD with the number of follow-up scans completed in the n=`r with(subset(df.bl, df.bl$NoScans > 1 & df.bl$subtype == "SPMS"), table(subtype))["SPMS"]` with >1 scan.
```{r}
with(subset(df.bl, df.bl$NoScans > 1 & df.bl$subtype == "SPMS"), cor.test(BrainPAD, NoScans, method = "spearman"))
```

## Longitudinal brain-predicted age trajectories
### Interaction between group and time MS vs. controls
Conditional growth model - random effects of participant and cohort
```{r}
model_int.group <- lmer(BrainPAD ~ control0rest1 * interval + poly(age_at_baseline_scan1, 2) + gender + field_strength + (interval|PatientID) + (1|Cohort), data = df, control = lmerControl(optimizer = "Nelder_Mead")) 
round(anova(model_int.group)["control0rest1:interval",],3)
```
#### Generate EMMs for healthy controls and for all MS/CIS combined, using annualised difference between baseline and final follow-up.
```{r message=FALSE, warning=FALSE}
fit <- (lmer(BrainPAD_change ~ control0rest1 + poly(age_at_baseline_scan1, 2) + gender + field_strength + (1|Cohort), data = z3))
emmeans(object = fit, ~ control0rest1)
```

### Interaction between group and time MS subtypes
Conditional growth model - random effects of participant and cohort
```{r}
model_int.group <- lmer(BrainPAD ~ subtype * interval + poly(age_at_baseline_scan1, 2) + gender + field_strength + (interval|PatientID) + (1|Cohort), data = subset(df, df$subtype != "control"), control = lmerControl(optimizer = "Nelder_Mead")) 
round(anova(model_int.group)["subtype:interval",],3)
```
#### Generate EMMs for healthy controls and for MS subtypes, using annualised difference between baseline and final follow-up.
```{r}
fit <- lmer(BrainPAD_change ~ subtype * interval + poly(age_at_baseline_scan1, 2) + gender + field_strength + (1|Cohort), data = subset(z3, z3$subtype != "control"), control = lmerControl(optimizer = "Nelder_Mead"))
emmeans(object = fit, ~ subtype)
```

### Slopes per group
Controls
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
control.slope <- summary(lmer(BrainPAD ~ interval + gender + age_at_baseline_scan1 + field_strength + (interval| PatientID) + (1|Cohort), data = subset(df, df$control0rest1 == "control")))$coef["interval","Estimate"]
control.intercept <- summary(lmer(BrainPAD ~ interval + gender + age_at_baseline_scan1 + field_strength + (interval| PatientID) + (1|Cohort), data = subset(df, df$control0rest1 == "control")))$coef["(Intercept)","Estimate"]
```
MS patients
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
ms.slope <- summary(lmer(BrainPAD ~ interval + gender + age_at_baseline_scan1 + (interval| PatientID) + (1|Cohort), data = subset(df, df$control0rest1 == "MS")))$coef["interval","Estimate"]
ms.intercept <- summary(lmer(BrainPAD ~ interval + gender + age_at_baseline_scan1 + (interval| PatientID) + (1|Cohort), data = subset(df, df$control0rest1 == "MS")))$coef["(Intercept)","Estimate"]
```

### Longitudinal brain-PAD by interval plots
```{r}
ggplot(data = df, aes(x = interval, y = BrainPAD, fill = subtype)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_line(aes(group = PatientID, colour = subtype), alpha = 0.3, linetype = 1, size = 0.25) +
  # geom_point(aes(fill = subtype), pch = 21, size = 0.5, alpha = 0.2) +
  geom_abline(slope = control.slope, intercept = control.intercept, col = "blue") +
  geom_abline(slope = ms.slope, intercept = ms.intercept, col = "red") +
  labs(x = "Time (years)", y = "Brain-predicted age difference (years)") +
  scale_fill_manual(values = ms.palette) +
  scale_color_manual(values = ms.palette) +
  theme_bw() + theme(legend.position = "none")

ggsave(filename = "~/Work/Brain ageing/Collaborations/MS/plots/longitudinal_brain-PAD_time_plot.pdf", height = 6, width = 6, useDingbats = FALSE)
```

```{r}
ggplot(data = df, aes(x = age_at_scan, y = BrainPAD, fill = subtype)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_line(aes(group = PatientID, colour = subtype), alpha = 0.5, linetype = 1, size = 0.25) +
  labs(x = "Time (years)", y = "Brain-predicted age difference (years)") +
  scale_fill_manual(values = ms.palette) +
  scale_color_manual(values = ms.palette) +
  facet_wrap(~ subtype, scales = "free_x") +
  theme_bw() + theme(legend.position = "none")
```
